### 2024-04-23 chap_RobPol/chap_RobPol.tex
RobPol + RL: Review by FC (part 1)

To deal with such uncertainties, decision-makers have several options: (i) resistance; (ii) resilience; (iii) static robustness; and (iv) adaptive robustness \cite{walker2012deep}. Where resistance consists in planning for the worst-case scenario, resilience aims at a fast recovery whatever the conditions in the future. In static robustness, one seeks for a roadmap that would perform ``satisfactorily'' in a wide range of plausible futures, whereas, a roadmap would be dynamically robust if it is prepared to adapt in case of a change in conditions. The adaptability of the policy was addressed in Chapter \ref{chap:chap_RL}. The objective of this chapter is to apply the method described in Section \ref{sec:meth:PCA} to deal with the static robustness of pathway roadmaps.  \citet{castrejon2020making} assessed policy mix folllowing the same philosophy of ``satisfactory level of performance'' as \cite{walker2012deep}. In their work, they mostly focused on the electricity sector, accounting for a variety of stakeholders and related interests using STET (Socio-Technical Energy Transition) models to capture more properly societal and behavioral aspects in relation with policy implementation, enriching purely techno-economic model, like EnergyScope, that usually assume rational choice within an overall cost minimization.  However, in the case of the transition pathway of a whole-energy system, the challenges stand here in the definition of the ``performance metric'' as well as the ``satisfactory level of performance''. Between the sole total transition cost and the entire set of installed technologies that give too few and too much information, respectively, the performance metric here is defined through the \gls{PCA} approach. Then, when comes the ``satisfactory level of performance'', we propose a relative level of performance through a comparative analysis of different roadmaps. In other terms, one roadmap will not be robust or not in itself but rather more or less robust than another one.

@@

\section{Definition of the principal components of the transition}
\label{sec:RobPol:PC_transition}
As detailed in Section \ref{subsec:meth:PCA:transition}, we have decided to define the directions of variation, \ie the robustness metrics, based on the installed capacities through the transition in the different end-use sectors, \ie electricity, \gls{HT} heat, \gls{LT} heat, passenger mobility, freight mobility, \gls{HVC}, ammonia and methanol. These capacities represent the technological roadmaps to supply these \gls{EUD} while respecting the \ce{CO2}-budget.  As introduced in Section \ref{subsec:meth:PCA:transition}, the data considered in this method come from the \gls{GSA} carried out on the perfect foresight optimisation of the Belgian transition pathway (see Chapter \ref{chap:atom_mol}). This gave 1260 different transitions resulting, for each of them, from the pathway optimisation subject to a sample of uncertain parameters (see Section \ref{subsec:uncert_charac}). Appendix \ref{app:UQ_tech_cap} gives the exhaustive distributions of the installed capacities among the different end-use sectors from the \gls{GSA}.

\subsection{Principal components of each representative year}
\label{subsec:RobPol:PC_year}
Before computing the \gls{PCs} of each representative year of the transition, we pre-preprocessed the raw data though scaling and managing the outliers detailed in Section \ref{subsec:meth:PCA:transition}. The second step discarded decentralised \gls{HP} running on methane. Without managing the outliers, this technology has a contribution in the \gls{PCs} of transition for several reasons. First of all, the \gls{LT} heating sector represents between 22\% and 27\% of the \gls{EUD} in the case of Belgium. Being part of this major sector, the scaling of decentralised gas \gls{HP} is higher than technologies in other sectors. Then, given the \gls{LP} implementation of EnergyScope Pathway, the gas heat pumps completely substitute electric heat pumps in specific conditions (less electricity and cheaper methane). However, theses cases represent less than 10\% of the 1260 runs. For these reasons, we have decided that managing the outliers was needed to assess the main directions of variation of the transition.

First of all, before investigating the \gls{PCs}, it is worth looking at the total variance of each representative year (see Table \ref{tab:design_variance}). Even though the absolute value of these variances has no physical meaning, we observe that the variations are more important at earlier stages of the transition. In other words, the further the transition goes, the more limited the degrees of freedom are to respect the \ce{CO2}-budget.

@@

\subsection{Principal components of the perfect foresight REF transitions}
\label{subsec:RobPol:PC_transition}
Based on the \gls{PCs} of each representative year, $\text{PC}_{y}$, the \gls{PCs} of the transition, \ie the metrics to assess the robustness of roadmaps, can be computed. Before aggregating and averaging similar $\text{PC}_{y}$, it is necessary to rank them to ensure capturing most of the transition variance in the subsequent $\text{PC}_{\text{transition}}$.

Based on the design variance captured by each $\text{PC}_{y}$ in their respective representative year, we can rank them (see Table \ref{tab:ranking_PCs}). Summing all of these variances over the different years results in a ``pseudo'' total variance of the transition. To construct the \gls{PCs} of the transition, $\text{PC}_{\text{transition}}$, we keep the $\text{PC}_{y}$ that captures at least 85\% of this total variance of the transition (see Section \ref{subsec:meth:PCA:transition}).  This results in keeping 14 $\text{PC}_{y}$: the first and second \gls{PCs} of each representative year and the third PC of 2025 and 2035.

@@

\newpage
\section{Robustness assessment of pathway roadmaps}
\label{sec:RobPol:Rob_Assessment}
Now that the performance metric is defined, we can assess the robustness of different roadmaps. These roadmaps are defined as the technological mix given by the deterministic optimisation of the perfect foresight pathway under certain conditions. The two first roadmaps are the \textbf{REF} and \textbf{SMR} cases (see Chapter \ref{chap:atom_mol}). All the uncertain parameters are considered at their nominal value in these two cases. The only difference consists in allowing \gls{SMR} from 2040 onward in the \textbf{SMR} case. On top of these two cases, we add a so-called ``robust'' case, \textbf{ROB}, in the same sense as in the works of \citet{bertsimas2004price} or \citet{Moret2017PhDThesis} accounting for a ``protection parameter''. Then, we assess the robustness of these three roadmaps via a similar ``three-step rolling horizon approach'' as in the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the variation of investments needed through myopic pathway optimisation under uncertainties and, (iii) projecting the strategies from the myopic optimisations on the robustness metrics defined via \gls{PCA}.

\subsection{A robust roadmap?}
\label{subsec:RobPol:Rob_roadmap}
Rather than considering all the uncertain parameters at their worst values as in the work of \citet{soyster1973convex}, we follow here the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}. In their works, the authors consider a factor $\Gamma_{\text{obj}}\in [0,d]$ that represents a ``protection parameter'' of the objective function where $d$ is the total number of uncertain parameters (see Section \ref{subsec:pce}). If $\Gamma_{\text{obj}}= 0$, where all the uncertain parameters are at their nominal value, we obtain the deterministic solution of the \textbf{REF} case. If $\Gamma_{\text{obj}}= d$, this means considering the ``fully robust'' solution where all the uncertain parameters are their worst value, as in \citet{soyster1973convex}.  Between these two extreme cases, the uncertain parameters to account for at their worst values follow the ranking given by the \gls{GSA} (see Chapter \ref{chap:atom_mol}). In practice, $\Gamma_{\text{obj}}= 1$ means considering the cost of purchasing electrofuels at its worst value. $\Gamma_{\text{obj}}= 2$ adds the industry \gls{EUD} at its worst value in the lot, and so on. In the present analysis, we consider the ``robust'' roadmap by setting $\Gamma_{\text{obj}}= 6$. Consequently, costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the interest rate and the variable OPEX of technologies are considered at their worst values, \ie the upper bound of their respective range. In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{HT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the interest rate is increased by 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.

@@

In the transport of freight, the ROB roadmap transitions from diesel to \gls{FC} trucks with the temporal use of electric trucks from 2025 to 2035. Similarly to the industrial electric heaters, it is a way to integrate the early installation of solar \gls{PV} and absorb part of its intermittent electricity. Freight transport via trains and boats are not different from the REF roadmap. About the passenger mobility, compared to the REF case, the shift from \gls{ICE} cars to \gls{BEV} occurs earlier in the transition, in line with the increased production of intermittent electricity from \gls{VRES}. \\

@@

In conclusion, the two main assumptions impacting the ROB roadmap strategy compared to the REF one were the higher cost of purchasing imported energy carriers (except electricity) and higher industrial \gls{EUD}. More expensive imported energy carriers lead to, on the one hand, an earlier full deployment of local \gls{VRES} (\ie \gls{PV}, wind onshore and offshore) supported by a higher electrification of the heating and transport sectors.  On the other hand, it increases the share of the total OPEX due to the consumption of resources (\ie excluding the OPEX of the technologies) in the total transition cost: \ie 56\% versus 38\% in the REF case. This, combined with the increase of industrial \gls{EUD}, encourages a more efficient use of the resources, \ie \gls{CHP} and \gls{HP}. The increased interest rate and variable OPEX of technologies have a more negligible impact on the roadmap strategy as they identically affect the entire set of technologies.

@@

No tested roadmap is more robust than the others on all the \gls{PCs} of transition (Table \ref{tab:projection}). The SMR roadmap spans over similar ranges as the REF case. The most significant difference concerns PC 5. This PC is driven by the variations in the \gls{HVC} sector. As detailed in Chapter \ref{chap:atom_mol}, one of the impacts of the installation of \gls{SMR} from 2040 onward is the delayed substitution of naphtha-crackers by \gls{MTO}. As naphtha-cracker is the more economical option in the near-term of a myopic transition, this smoother modal shift leads to less variations of installed capacities. This makes the SMR roadmap more robust than the REF case.

@@

The ROB roadmap is more robust than the REF case for all PCs but PC 4. About PC 1, it is driven by the variation of installed capacities of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors. In the ROB case, the earlier full deployment of solar \gls{PV} makes this roadmap 49\% more robust towards these variations. For the same reasons, the ROB roadmap is 37\% more robust for PC 2. The level of robustness is smaller than for PC 1 as the ROB roadmap substitutes sooner its decentralised gas boilers. In myopic transitions though, this technology keeps on being installed in the mid-term. Concerning PC 3, this higher robustness is explained by variations in the installed capacities of diesel trucks, biomass-to-methanol and ammonia-\gls{CCGT} that are 32\%, 30\% and 15\% smaller than in the REF case, respectively.  Finally, the ROB roadmap is 21\% less robust for PC 5. Even though this roadmap accounts for the additional capacities of \gls{MTO} in the long-term to supply the potential additional demand of \gls{HVC}, the near-term capacities of naphtha-crackers are smaller. As aforementioned, this solution is favoured in the myopic optimisation which makes the ROB roadmap less robust to these variations.

\begin{table}[htbp!]
\caption{Comparison of the robustness with the REF case where robustness is defined by the width of the 95\% confidence range of the additional installed capacities projected on the PC. Differences below 5\% are not shown ($\simeq$).}
\label{tab:projection}
\centering
\begin{tabular}{c| c c}
\toprule
\multirow{3}{*}{\textbf{PC}} & \multicolumn{2}{c}{\textbf{Robustness comparison versus REF}}\\
& SMR & ROB \\
\midrule
1 & $\simeq$ & +49\% \\
2 &  $\simeq$ & +37\%\\
3 &  $\simeq$ & +33\%\\
4 & $\simeq$ & $\simeq$ \\
5 &  +34\% & -21\%\\
\bottomrule							
\end{tabular}
\end{table}

@@

Then, we have tested three roadmaps resulting from different perfect foresight optimisations: REF, SMR and ROB. Where the first two have been broadly discussed in Chapter \ref{chap:atom_mol}, the ``robust'' (ROB) case was considering the worst value of the uncertain parameters impacting the most the total transition cost following the same approach as \citet{Moret2017PhDThesis}. From these roadmaps setting the minimal installed capacities over the transition, we have assessed the variation the additional capacities needed in myopic optimisations subject to uncertainties. This showed that counting on the installation of \gls{SMR} by 2040 was providing the same level of robustness as the reference case in the key directions of variation. However, the robust (ROB) investment roadmap --- which, in our case, anticipates the full deployment of local \gls{VRES} and the integration of efficient technologies --- is more robust than the reference scenario. Even though this roadmap represents 2.6\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors (49\%) and technologies in the transport of freight (33\%).

>>>>

To deal with such uncertainties, decision-makers have several options: (i) resistance; (ii) resilience; (iii) static robustness; and (iv) adaptive robustness \cite{walker2012deep}. Where resistance consists in planning for the worst-case scenario, resilience aims at a fast recovery whatever the conditions in the future. In static robustness, one seeks for a roadmap that would perform ``satisfactorily'' in a wide range of plausible futures, whereas, a roadmap would be dynamically robust if it is prepared to adapt in case of a change in conditions. The adaptability of the policy was addressed in Chapter \ref{chap:chap_RL}. The objective of this chapter is to apply the method described in Section \ref{sec:meth:PCA} to deal with the static robustness of pathway roadmaps.  \citet{castrejon2020making} assessed policy mix folllowing the same philosophy of ``satisfactory level of performance'' as \cite{walker2012deep}. In their work, they mostly focused on the electricity sector, accounting for a variety of stakeholders and related interests using STET (Socio-Technical Energy Transition) models to capture more properly societal and behavioral aspects in relation with policy implementation, enriching purely techno-economic model, like EnergyScope, that usually assume rational choice within an overall cost minimization.  However, in the case of the transition pathway of a whole-energy system, the challenges stand here in the definition of the ``performance metric'' as well as the ``satisfactory level of performance''. Between the total transition cost giving too few information and the entire set of installed technologies giving too much, the performance metric here is defined through the \gls{PCA} approach. Then, when comes the ``satisfactory level of performance'', we propose a relative level of performance through a comparative analysis of different roadmaps. Each roadmap will be ranked in terms of robustness.

@@

\section{Definition of the principal components of the transition}
\label{sec:RobPol:PC_transition}
As detailed in Section \ref{subsec:meth:PCA:transition}, we have decided to define the directions of variation, \ie the robustness metrics, based on the installed capacities through the transition in the different end-use sectors, \ie electricity, \gls{HT} heat, \gls{LT} heat, passenger mobility, freight mobility, \gls{HVC}, ammonia and methanol. These capacities represent the technological roadmaps to supply these \gls{EUD} while respecting the \ce{CO2}-budget.  Data considered in this method come from the \gls{GSA} carried out on the perfect foresight optimisation of the Belgian transition pathway (see Chapter \ref{chap:atom_mol}). This gave 1260 different transitions resulting, for each of them, from the pathway optimisation subject to a sample of uncertain parameters (see Section \ref{subsec:uncert_charac}). Appendix \ref{app:UQ_tech_cap} gives the exhaustive distributions of the installed capacities among the different end-use sectors from the \gls{GSA}.

\subsection{Principal components of each representative year}
\label{subsec:RobPol:PC_year}
Before computing the \gls{PCs} of each representative year, we pre-preprocessed the raw data through scaling and managing the outliers (see Section \ref{subsec:meth:PCA:transition}). Managing the outliers is needed when assessing the main directions of variation of the transition. For example, decentralised \gls{HP} running on methane is discarded. This technology would have a contribution in the \gls{PCs} of transition for several reasons. First of all, the \gls{LT} heating sector represents between 22\% and 27\% of the \gls{EUD} in the case of Belgium. Being part of this major sector, the scaling of decentralised gas \gls{HP} is higher than technologies in other sectors. Then, given the \gls{LP} implementation of EnergyScope Pathway, the gas heat pumps completely substitute electric heat pumps in specific conditions (less electricity and cheaper methane). However, theses cases represent less than 10\% of the 1260 runs. For these reasons, we have decided that managing the outliers was needed to assess the main directions of variation of the transition.

The total variance decreases by more than 50\% between the earlier stages and the final ones (see Table \ref{tab:design_variance}). Even though the absolute value of these variances has no physical meaning, we observe that the variations are more important at earlier stages of the transition. In other words, the further the transition goes, the more limited the degrees of freedom are to respect the \ce{CO2}-budget.

@@

\subsection{Principal components of the perfect foresight REF transitions}
\label{subsec:RobPol:PC_transition}
Based on the \gls{PCs} of each representative year (34 in total), this section identifies a sufficient subset to characterise the roadmaps (and assess their robustness in Section \ref{subsec:RobPol:Projection}).

Before aggregating and averaging similar $\text{PC}_{y}$, it is necessary to rank them to ensure capturing most of the transition variance in the subsequent $\text{PC}_{\text{transition}}$.This ranking is based on the design variance captured by each $\text{PC}_{y}$ in their respective representative year (see Table \ref{tab:ranking_PCs}). Summing all of these variances over the different years results in a ``pseudo'' total variance of the transition. To construct the \gls{PCs} of the transition, $\text{PC}_{\text{transition}}$, we keep the $\text{PC}_{y}$ that captures at least 85\% of this total variance of the transition (see Section \ref{subsec:meth:PCA:transition}).  This results in keeping 14 $\text{PC}_{y}$: the first and second \gls{PCs} of each representative year and the third PC of 2025 and 2035.

@@

\newpage
\section{Robustness assessment of pathway roadmaps}
\label{sec:RobPol:Rob_Assessment}
Now that the performance metric is defined, we can assess the robustness of different roadmaps. These roadmaps are defined as the technological mix given by the deterministic optimisation of the perfect foresight pathway under certain conditions. The two first roadmaps are the \textbf{REF} and \textbf{SMR} cases (see Chapter \ref{chap:atom_mol}). All the uncertain parameters are considered at their nominal value in these two cases. The only difference consists in allowing \gls{SMR} from 2040 onward in the \textbf{SMR} case. On top of these two cases, we add a so-called ``robust'' case, \textbf{ROB}, in the same sense as in the works of \citet{bertsimas2004price} or \citet{Moret2017PhDThesis} accounting for a ``protection parameter''. Then, we assess the robustness of these three roadmaps via a ``three-step rolling horizon approach'' extended from the work of \citet{moret2020overcapacity}: (i) setting the initial investment strategies provided by the roadmaps and, (ii) assessing the variation of investments needed through myopic pathway optimisation under uncertainties and, (iii) projecting the strategies from the myopic optimisations on the robustness metrics defined via \gls{PCA}.

\subsection{A robust roadmap?}
\label{subsec:RobPol:Rob_roadmap}
Rather than considering all the uncertain parameters at their worst values as in the work of \citet{soyster1973convex}, we follow here the robust approach of \citet{bertsimas2004price} and \citet{Moret2017PhDThesis}. In their works, the authors consider a factor $\Gamma_{\text{obj}}\in [0,d]$ that represents a ``protection parameter'' of the objective function where $d$ is the total number of uncertain parameters (see Section \ref{subsec:pce}). If $\Gamma_{\text{obj}}= 0$, where all the uncertain parameters are at their nominal value, we obtain the deterministic solution of the \textbf{REF} case. If $\Gamma_{\text{obj}}= d$, this means considering the ``fully robust'' solution where all the uncertain parameters are their worst value, as in \citet{soyster1973convex}.  Between these two extreme cases, the uncertain parameters to account for at their worst values follow the ranking given by the \gls{GSA} (see Chapter \ref{chap:atom_mol}). In practice, $\Gamma_{\text{obj}}= 1$ means considering the cost of purchasing electrofuels at its worst value. $\Gamma_{\text{obj}}= 2$ adds the industry \gls{EUD} at its worst value, and so on. In the present analysis, we label the roadmap as ``robust'' (ROB) by setting $\Gamma_{\text{obj}}= 6$. Consequently, the following six parameters are set at their worst value (upper bound of their range): costs of purchasing electrofuels, fossil fuels and biofuels as well as the industry \gls{EUD}, the interest rate and the variable OPEX of technologies. In practice, this means that, in the ROB case, imported energy carriers, except electricity, are 179.8\% more expensive, the industrial demand is, by 2050, 23.9\% higher\footnote{As detailed in Chapter \ref{chap:case_study}, the industrial demand encompasses the whole \gls{NED} and \gls{HT} heat demand, as well as the industrial share of electricity and \gls{LT} heat demands.}, the interest rate is increased by 46.2\% (\ie 2.2\% versus 1.5\% in the REF case) and the variable OPEX of technologies, $c_{\text{\emph{maint}}}$, is 35.7\% higher.

@@

In the ROB roadmap, the transport of freight by trucks first transitions from diese to electric trucks and then, after 2035, to \gls{FC} trucks. Used as a way to support the the early installation of solar \gls{PV}, electric trucks are then replaced by \gls{FC} ones given the limited availability of renewable electricity and its more cost-effective use in other sectors such as private mobility. Freight transport via trains and boats are not different from the REF roadmap. About the passenger mobility, compared to the REF case, the shift from \gls{ICE} cars to \gls{BEV} occurs earlier in the transition, in line with the increased production of intermittent electricity from \gls{VRES}. \\

@@

In conclusion, the two main assumptions impacting the ROB roadmap strategy compared to the REF one were the higher cost of purchasing imported energy carriers (except electricity) and higher industrial \gls{EUD}. More expensive imported energy carriers lead to an earlier full deployment of local \gls{VRES} (\ie \gls{PV}, wind onshore and offshore) and an increase of the share of the total OPEX due to the consumption of resources (\ie excluding the OPEX of the technologies) in the total transition cost: \ie 56\% versus 38\% in the REF case. This, combined with the increase of industrial \gls{EUD}, encourages a more efficient use of the resources, \ie \gls{CHP} and \gls{HP}. The increased interest rate and variable OPEX of technologies have a more negligible impact on the roadmap strategy as they identically affect the entire set of technologies.

@@

The SMR roadmap spans over similar ranges as the REF case (see Table \ref{tab:projection}). The most significant difference concerns PC 5. This PC is driven by the variations in the \gls{HVC} sector. As detailed in Chapter \ref{chap:atom_mol}, one of the impacts of the installation of \gls{SMR} from 2040 onward is the delayed substitution of naphtha-crackers by \gls{MTO}. As naphtha-cracker is the more economical option in the near-term of a myopic transition, this smoother modal shift leads to less variations of installed capacities. This makes the SMR roadmap more robust than the REF case.

@@

The ROB roadmap is more robust than the REF case for the first three PCs. About PC\,1, it is driven by the variation of installed capacities of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors. In the ROB case, the earlier full deployment of solar \gls{PV} makes this roadmap 49\% more robust towards these variations. For the same reasons, the ROB roadmap is 37\% more robust for PC\,2. The level of robustness is smaller than for PC 1 as the ROB roadmap substitutes sooner its decentralised gas boilers. In myopic transitions though, this technology keeps on being installed in the mid-term. Concerning PC\,3, this higher robustness is explained by variations in the installed capacities of diesel trucks, biomass-to-methanol and ammonia-\gls{CCGT} that are 32\%, 30\% and 15\% smaller than in the REF case, respectively.  Finally, the ROB roadmap is 21\% less robust for PC\,5. Even though this roadmap accounts for the additional capacities of \gls{MTO} in the long-term to supply the potential additional demand of \gls{HVC}, the near-term capacities of naphtha-crackers are smaller. As aforementioned, this solution is favoured in the myopic optimisation which makes the ROB roadmap less robust to these variations.

\begin{table}[htbp!]
\caption{Comparison of the robustness with the REF case where robustness is defined by the width of the 95\% confidence range of the additional installed capacities projected on the PC. Differences below 5\% are not shown ($\simeq$).}
\label{tab:projection}
\centering
\begin{tabular}{c l| c c}
\toprule
\multirow{2}{*}{\textbf{PC}} & \multirow{2}{*}{\textbf{Main contributors}} & \multicolumn{2}{c}{\textbf{Robustness comparison versus REF}}\\
& & SMR & ROB \\
\midrule
1 & \gls{PV} and electrification & $\simeq$ & +49\% \\
2 & \gls{PV} and dec.heating & $\simeq$ & +37\%\\
3 & Freight transport & $\simeq$ & +33\%\\
4 & Dec. oil boilers & $\simeq$ & $\simeq$ \\
5 &  Naphtha-crackers and \gls{MTO} & +34\% & -21\%\\
\bottomrule							
\end{tabular}
\end{table}

@@

Then, we have tested three roadmaps resulting from different perfect foresight optimisations: REF, SMR and ROB. Where the first two have been broadly discussed in Chapter \ref{chap:atom_mol}, the ``robust'' (ROB) case was considering the worst value of the uncertain parameters impacting the most the total transition cost following the same approach as \citet{Moret2017PhDThesis}. From these roadmaps setting the minimal installed capacities over the transition, we have assessed the variation of the additional capacities needed in myopic optimisations subject to uncertainties. This showed that counting on the installation of \gls{SMR} by 2040 was providing the same level of robustness as the reference case in the key directions of variation. However, the robust (ROB) investment roadmap --- which, in our case, anticipates the full deployment of local \gls{VRES} and the integration of efficient technologies --- is more robust than the reference scenario. Even though this roadmap represents 2.6\% more cumulative \gls{CAPEX} over the transition than the REF case, it provides more robustness towards the variations of \gls{PV} supported by the electrification of \gls{HT} and \gls{LT} heating sectors (49\%) and technologies in the transport of freight (33\%).

### 2024-04-23 chap_case_study/chap_case_study.tex
RL chapter: submission for review
RobPol + RL: Review by FC (part 1)

Third, where previous works considered a prescribed \ce{CO2} trajectory to reach carbon-neutrality by 2050 \cite{limpens2021generating, limpens2024pathway}, the case study analysed in this thesis is subject to a \ce{CO2}-budget for the transition, \ie limiting the total amount of emissions over the transition. Based on the estimated world budget provided by the \gls{IPCC} to limit the global warming to +1.5°C by the end of the century, the grandfathering approach has been used to allocate part of this budget to the Belgian energy transition.

@@

Then, besides their costs, the resources are either limited or unlimited in terms of availability and either renewable or not. The limitation in terms of availability can be direct or indirect. On the one hand, woody (23.4\,TWh) and wet biomass (38.9\,TWh) are directly limited by their local potentials and the consumption of waste (17.8\,TWh) and coal (33.4\,TWh) is assumed not to exceed the current use. The local potential of biomass, \ie about 60\,TWh in total, could be considered as optimistic given the work of \citet{colla2024navigating} rather limiting this potential to about 40\,TWh. On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and the choice to limit nuclear power plants to 6\,GW, respectively. In line with the work of \citet{PATHS2050} and the maximum capacity of conventional nuclear reactors that have been installed in Belgium, the same 6\,GW are assumed to be the maximum capacity for \gls{SMR}. Imported electricity is limited in two ways: the potential of instantaneous capacity of interconnection with neighbouring countries (\ie 11.9\,GW by 2050 \cite{ELIA_2050}) and a limitation to 30\% of the yearly electricity end-use demand (\ie 32.4\,TWh by 2050) \cite{limpens2021generating}.

>>>>

Third, where previous works considered a prescribed \ce{CO2} trajectory to reach carbon-neutrality by 2050 \cite{limpens2021generating, limpens2024pathway}, the case study analysed in this thesis is subject to a \ce{CO2}-budget for the transition, \ie limiting the total amount of emissions over the transition. Based on the estimated world budget provided by the \gls{IPCC} to limit the global warming to +1.5°C above pre-industrial levels by the end of the century, the grandfathering approach has been used to allocate part of this budget to the Belgian energy transition.

@@

Then, besides their costs, the resources are either limited or unlimited in terms of availability and either renewable or not. The limitation in terms of availability can be direct or indirect. On the one hand, woody (23.4\,TWh) and wet biomass (38.9\,TWh) are directly limited by their local potentials and the consumption of waste (17.8\,TWh) and coal (33.4\,TWh) is assumed not to exceed the current use. Similarly to \citet{limpens2021generating}, we constrained the endogenous non-renewable wastes to be consumed locally. The local potential of biomass, \ie about 60\,TWh in total, could be considered as optimistic given the work of \citet{colla2024navigating} rather limiting this potential to about 40\,TWh. On the other hand, wind, solar, hydro and uranium are limited by the technical potentials of \gls{PV} panels (59.2\,GW), onshore (10\,GW) and offshore (6\,GW) wind turbines, run-of-the-river power plants (0.1\,GW) and the choice to limit nuclear power plants to 6\,GW, respectively. In line with the work of \citet{PATHS2050} and the maximum capacity of conventional nuclear reactors that have been installed in Belgium, the same 6\,GW are assumed to be the maximum capacity for \gls{SMR}. Imported electricity is limited in two ways: the potential of instantaneous capacity of interconnection with neighbouring countries (\ie 11.9\,GW by 2050 \cite{ELIA_2050}) and a limitation to 30\% of the yearly electricity end-use demand (\ie 32.4\,TWh by 2050) \cite{limpens2021generating}.

### 2024-04-23 chap_intro_ccl/conclusion.tex
RobPol + RL: Review by FC (part 1)

Regarding the \gls{RL} approach, two threads of future development have been identified: reward shaping and multi-fidelity.  Rather than giving a sparse reward at the end of the episode, reward shaping provides additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights \cite{hu2020learning}. Reward shaping aims at accelerating the learning or guiding the agent towards achieving the desired behaviour more efficiently. In our case, the agent receives a sparse reward signal, indicating success or failure at the end of an episode. However, this sparse reward signal may not provide enough information for the agent to learn effectively, leading to slow convergence or difficulty in learning the optimal policy. Reward shaping addresses this issue. These intermediate rewards can help guide the agent towards desirable states or actions, making the learning process more efficient and effective. However, reward shaping should be applied with caution, as poorly designed reward functions can lead to unintended consequences such as suboptimal policies, reward hacking, or overfitting to the shaped rewards rather than learning the underlying task.

>>>>

Regarding the \gls{RL} approach, two threads of future development have been identified: reward shaping and multi-fidelity.  First, the reward function could be adapted to incentivise more the successful transitions. Then, rather than giving a sparse reward at the end of the episode, reward shaping provides additional, intermediate rewards during the learning process based on various heuristics, domain knowledge, or problem-specific insights \cite{hu2020learning}. Reward shaping aims at accelerating the learning or guiding the agent towards achieving the desired behaviour more efficiently. In our case, the agent receives a sparse reward signal, indicating success or failure at the end of an episode. However, this sparse reward signal may not provide enough information for the agent to learn effectively, leading to slow convergence or difficulty in learning the optimal policy. Reward shaping addresses this issue. These intermediate rewards can help guide the agent towards desirable states or actions, making the learning process more efficient and effective. However, reward shaping should be applied with caution, as poorly designed reward functions can lead to unintended consequences such as suboptimal policies, reward hacking, or overfitting to the shaped rewards rather than learning the underlying task.
